Follow-up Samples

### 2017-05-03 submission/manuscript.Rmd
update manuscirpt with new supplemental figs

***The effects of treatment are not consistent across treatment groups.***
We used two approaches to identify those bacterial populations that change between the two samples for each diagnosis group. First, we sought to identify individual OTUs that could account for the change in overall community structure. However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups [Figure S1]. It is likely that high inter-individual variation and the irregular distribution of OTUs across individuals limited the statistical power of the test. To overcome these problems we developed Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups. Because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs to limit the likelihood that the models would overfit the data. Despite this restriction, the models performed well (adenoma AUC range = `r format(min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`). Interestingly, the 10 OTUs that were used for each model had little overlap with each other [Figure 2]. These results support the earlier community-wide analysis where we observed that the treatment had an impact on the overall community structure; however, the effect of treatment was not consistent across patients and diagnosis groups.

***Post-treatment samples from patients with carcinoma more closely resemble those of a normal colon.***
Next, we determined whether treatment changed the microbiota in a way that the post-treatment communities resembled that of patients with normal colons. To test this, we used an expanded cohort of `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals that were diagnosed under the same protocol as having normal colons or colons with adenoma, advanced adenoma, or carcinoma [Table 2]. We then constructed Random Forest models to classify the study samples, with the 3 diagnosis groups (adenoma, advanced adenoma, or carcinoma), or having a normal colon. The models performed well (adenoma AUC range =`r format(min(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)`; Figure S2). The OTUs that were incorporated into the adenoma and advanced adenoma models largely overlapped and those OTUs that were used to classify the carcinoma samples were largely distinct from those of the other two models [Figure 3A]. Among the OTUs that were shared across the three models were those populations generally considered beneficial to their host (e.g. *Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, *Dorea*, *Anaerostipes*, and *Roseburia*) [Figures 3B]. Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014] [Figure S3] with some individuals showing are marked decrease in relative abundance [Figure S4]. Finally, we applied these three models to the pre and post-treatment samples for each diagnosis group and quantified the change in the positive probability of the model. A decrease in the positive probability would indicate that the microbiota more closely resembled that of a patient with a normal colon. There was no significant change in the positive probability for the adenoma or advanced adenoma groups [Figure 4]. The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased with treatment, suggesting a shift toward a normal microbiota for most individuals. Only, `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the 26 patients (`r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`%) who were diagnosed with a carcinoma had a higher positive probability after treatment; one of those was re-diagnosed with carcinoma on the follow up visit. These results indicate that, although there were changes in the microbiota associated with treatment, those experienced by  patients with carcinoma after treatment yielded gut bacterial communities of greater similarity to that of a normal colon.

@@

Patients diagnosed with adenoma and advanced adenoma, however, did not experience a shift towards a community structure that resembled those with normal colons. This may be due to the fundamental differences between the features of adenomas and advanced adenomas and carcinoma. Specifically, carcinomas may create an inflammatory milieu that would impact the structure of the community and removal of that stimulus would alter said structure. In addition, it is possible that the difference between the microbiota of patients with adenoma and advanced adenoma and those with normal colons is subtle. This is supported by the reduced ability of our models to correctly classify patients with adenomas and advanced adenomas relative to those diagnosed with carcinomas [Figure S2]. Given the irregular distribution of microbiota across patients in the different diagnosis groups, it is possible that we lacked the statistical power to adequately characterize the change in the communities following treatment.

@@

**Figure S1: Distribution of P-values from Paired Wilcoxson Analysis of All OTUs Before and After Treatment**

**Figure S2: ROC Curves of the Adenoma, Advanced Adenoma, and Carcinoma Models.** A) Adenoma ROC curve: The light green shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification. B) Advanced Adenoma ROC curve: The light yellow shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark yellow line represents the model using 100% of the data set and what was used for subsequent classification. C) Carcinoma ROC curve: The light red shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark red line represents the model using 100% of the data set and what was used for subsequent classification.

**Figure S3: Summary of Important OTUs for the Adenoma, Advanced Adenoma, and Carcinoma Models.** A) MDA of the most important variables in the adenoma model. The dark green point represents the mean and the lighter green points are the value of each of the 100 different runs. B) Summary of Important Variables in the advanced adenoma model. MDA of the most important variables in the SRN model. The dark yellow point represents the mean and the lighter yellow points are the value of each of the 100 different runs. C) MDA of the most important variables in the carcinoma model. The dark red point represents the mean and the lighter red points are the value of each of the 100 different runs.

>>>>

***The effects of treatment are not consistent across treatment groups.***
We used two approaches to identify those bacterial populations that change between the two samples for each diagnosis group. First, we sought to identify individual OTUs that could account for the change in overall community structure. However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups (P-value `r ifelse(all_OTU_pair_wilcox > 0.05, invisible(">"), invisible("<"))` 0.05). It is likely that high inter-individual variation and the irregular distribution of OTUs across individuals limited the statistical power of the test. To overcome these problems we developed Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups. Because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs to limit the likelihood that the models would overfit the data. Despite this restriction, the models performed well (adenoma AUC range = `r format(min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`). Interestingly, the 10 OTUs that were used for each model had little overlap with each other [Figure 2]. These results support the earlier community-wide analysis where we observed that the treatment had an impact on the overall community structure; however, the effect of treatment was not consistent across patients and diagnosis groups.

***Post-treatment samples from patients with carcinoma more closely resemble those of a normal colon.***
Next, we determined whether treatment changed the microbiota in a way that the post-treatment communities resembled that of patients with normal colons. To test this, we used an expanded cohort of `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals that were diagnosed under the same protocol as having normal colons or colons with adenoma, advanced adenoma, or carcinoma [Table 2]. We then constructed Random Forest models to classify the study samples, with the 3 diagnosis groups (adenoma, advanced adenoma, or carcinoma), or having a normal colon. The models performed well (adenoma AUC range =`r format(min(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)`; Figure S1). The OTUs that were incorporated into the adenoma and advanced adenoma models largely overlapped and those OTUs that were used to classify the carcinoma samples were largely distinct from those of the other two models [Figure 3A]. Among the OTUs that were shared across the three models were those populations generally considered beneficial to their host (e.g. *Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, *Dorea*, *Anaerostipes*, and *Roseburia*) [Figures 3B]. Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014] [Figure S2] with some individuals showing are marked decrease in relative abundance [Figure S3]. Finally, we applied these three models to the pre and post-treatment samples for each diagnosis group and quantified the change in the positive probability of the model. A decrease in the positive probability would indicate that the microbiota more closely resembled that of a patient with a normal colon. There was no significant change in the positive probability for the adenoma or advanced adenoma groups [Figure 4]. The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased with treatment, suggesting a shift toward a normal microbiota for most individuals. Only, `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the 26 patients (`r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`%) who were diagnosed with a carcinoma had a higher positive probability after treatment; one of those was re-diagnosed with carcinoma on the follow up visit. These results indicate that, although there were changes in the microbiota associated with treatment, those experienced by  patients with carcinoma after treatment yielded gut bacterial communities of greater similarity to that of a normal colon.

@@

Patients diagnosed with adenoma and advanced adenoma, however, did not experience a shift towards a community structure that resembled those with normal colons. This may be due to the fundamental differences between the features of adenomas and advanced adenomas and carcinoma. Specifically, carcinomas may create an inflammatory milieu that would impact the structure of the community and removal of that stimulus would alter said structure. In addition, it is possible that the difference between the microbiota of patients with adenoma and advanced adenoma and those with normal colons is subtle. This is supported by the reduced ability of our models to correctly classify patients with adenomas and advanced adenomas relative to those diagnosed with carcinomas [Figure S1]. Given the irregular distribution of microbiota across patients in the different diagnosis groups, it is possible that we lacked the statistical power to adequately characterize the change in the communities following treatment.

@@

**Figure S1: ROC Curves of the Adenoma, Advanced Adenoma, and Carcinoma Models.** A) Adenoma ROC curve: The light green shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification. B) Advanced Adenoma ROC curve: The light yellow shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark yellow line represents the model using 100% of the data set and what was used for subsequent classification. C) Carcinoma ROC curve: The light red shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark red line represents the model using 100% of the data set and what was used for subsequent classification.

**Figure S2: Summary of Important OTUs for the Adenoma, Advanced Adenoma, and Carcinoma Models.** A) MDA of the most important variables in the adenoma model. The dark green point represents the mean and the lighter green points are the value of each of the 100 different runs. B) Summary of Important Variables in the advanced adenoma model. MDA of the most important variables in the SRN model. The dark yellow point represents the mean and the lighter yellow points are the value of each of the 100 different runs. C) MDA of the most important variables in the carcinoma model. The dark red point represents the mean and the lighter red points are the value of each of the 100 different runs.

**Figure S3: Pre and Post Treatment Relative Abundance of CRC associated OTUs within the Carcinoma Model.**
